---
title: "PCA analysis"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.retina = 2)
library(tidyverse)
library(ggpubr)
library(ggtree)
library(phytools)
library(ggrepel)
source("scripts/my_color.R")
```


### PCA and admixture analysis

We firstly used pcangsd to calculate covariance of all SNPs and plotted the first two principles. The strongest variance was observed between samples from MI and all the other north reefs. The samples from north reefs seems form a big cluster without any clustering pattern with only internal genetic diversity reflected along PC2. We also identified 4 potential hybrids, consistently in PCA plot and Admixture plot. These hybrids were excluded in selection analysis, IBS tree, and demographic modelling.

```bash
pcangsd --beagle atenuis.beagle.gz --threads 48 --admix --admix_auto 10000 --out atenuis.pcangsd
```

```{r}
covmat <- read_table("data/hpc/pca/atenius.ind212.unique_mdust1M.pcangsd.cov",col_names = F) %>% as.matrix()

sample_ids <- read_tsv("data/hpc/pca/ind212.txt", col_names = "sample_id",show_col_types = FALSE) %>% pull(sample_id)

colnames(covmat)<- sample_ids
rownames(covmat) <- sample_ids

pop_eigen <- eigen(covmat)
sample_table <- read_csv("data/hpc/qc/summary_data.csv",show_col_types = FALSE) %>% select(sample_id,pop,location,reef)

pop_pca12 <- data.frame(x=pop_eigen$vectors[,1],y=pop_eigen$vectors[,2],sample_id = sample_ids) %>% left_join(sample_table) %>% mutate(pop_order=site_order()[pop])
pop_pca23 <- data.frame(x=pop_eigen$vectors[,2],y=pop_eigen$vectors[,3],sample_id = sample_ids) %>% left_join(sample_table) %>% mutate(pop_order=site_order()[pop])
pc1 <- round(((pop_eigen$values)/sum(pop_eigen$values))*100,2)[1]
pc2 <- round(((pop_eigen$values)/sum(pop_eigen$values))*100,2)[2]
pc3 <- round(((pop_eigen$values)/sum(pop_eigen$values))*100,2)[3]

ggplot(pop_pca12, aes(x=x,y=y)) + geom_point(aes(color=reorder(pop,pop_order),shape=reef),size=2)+ 
  geom_text_repel(data=pop_pca12 %>% filter(sample_id %in% c("MI-1-1_S10","PI-1-16_S16","DI-2-4_S17","ARL_15_S69")),aes(x=x,y=y,label=sample_id),size=2,hjust=2,vjust=1) + 
  geom_point(data=pop_pca12 %>% filter(sample_id %in% c("MI-1-1_S10","PI-1-16_S16","DI-2-4_S17","ARL_15_S69")),aes(x=x,y=y),size=.8,color="black") +
  scale_color_manual(values = site_colors(), labels=site_labels()) + 
  theme_test() + labs(x=paste0("PC1 (",pc1,"%)"),y=paste0("PC2 (",pc2,"%)"), shape="Location", color="Reef")

ggplot(pop_pca23, aes(x=x,y=y)) + geom_point(aes(color=reorder(pop,pop_order),shape=reef),size=2)+
  scale_color_manual(values = site_colors(), labels=site_labels()) + 
  theme_test() + labs(x=paste0("PC2 (",pc2,"%)"),y=paste0("PC3 (",pc3,"%)"), shape="Location", color="Reef")

hybrids <-c("MI-1-1_S10","PI-1-16_S16","DI-2-4_S17","ARL_15_S69")
```

```{r}
q2_admixture <- read_table("data/hpc/pca/atenius.ind212.unique_mdust1M.pcangsd.admix.2.Q", col_names = c("X2","X1")) %>%
  add_column(sample_id = sample_ids) %>% 
  gather(cluster, proportion, -sample_id) %>% 
  left_join(sample_table) %>% 
  mutate(pop_order = site_order()[pop])

p1 <- ggplot(q2_admixture ,aes(x=reorder(sample_id,pop_order),y=proportion)) + 
  geom_bar(aes(fill=cluster,color=cluster),stat="identity") + 
  geom_point(data=q2_admixture %>% filter(sample_id %in% hybrids),aes(x=reorder(sample_id,pop_order)),y=1,shape=8) + 
  facet_grid(reorder(location,desc(pop_order))~., scales = "free_y") + coord_flip() + theme_pubclean() + labs(x="",y="Proportion (K=2)") + 
  scale_fill_brewer(palette = "Set2") + 
  scale_color_brewer(palette = "Set2") + 
theme(axis.text.y = element_blank(),
        legend.position = "none", 
        axis.ticks = element_blank(),
        panel.border = element_blank(), 
        panel.background = element_blank(),
        panel.grid.minor = element_blank(),
        strip.text = element_text(size=12))
p1
```


### LD-Pruned results

```{r}
covmat_unlinked <- read_table("data/hpc/pca/atenius.ld_pruned_snps.pcangsd.cov",col_names = F) %>% as.matrix()

colnames(covmat_unlinked)<- sample_ids
rownames(covmat_unlinked) <- sample_ids

pop_eigen_unlinked <- eigen(covmat_unlinked)

pop_pca12_unlinked <- data.frame(x=pop_eigen_unlinked$vectors[,1],y=pop_eigen_unlinked$vectors[,2],sample_id = sample_ids) %>% left_join(sample_table) %>% mutate(pop_order=site_order()[pop])
pop_pca23_unlinked <- data.frame(x=pop_eigen_unlinked$vectors[,2],y=pop_eigen_unlinked$vectors[,3],sample_id = sample_ids) %>% left_join(sample_table) %>% mutate(pop_order=site_order()[pop])
pc1 <- round(((pop_eigen_unlinked$values)/sum(pop_eigen$values))*100,2)[1]
pc2 <- round(((pop_eigen_unlinked$values)/sum(pop_eigen$values))*100,2)[2]
pc3 <- round(((pop_eigen_unlinked$values)/sum(pop_eigen$values))*100,2)[3]

ggplot(pop_pca12_unlinked, aes(x=x,y=y)) + geom_point(aes(color=reorder(pop,pop_order),shape=reef),size=2)+
  scale_color_manual(values = site_colors(), labels=site_labels()) + 
  theme_test() + labs(x=paste0("PC1 (",pc1,"%)"),y=paste0("PC2 (",pc2,"%)"), shape="Location", color="Reef")

ggplot(pop_pca23_unlinked, aes(x=x,y=y)) + geom_point(aes(color=reorder(pop,pop_order),shape=reef),size=2)+
  scale_color_manual(values = site_colors(), labels=site_labels()) + 
  theme_test() + labs(x=paste0("PC2 (",pc2,"%)"),y=paste0("PC3 (",pc3,"%)"), shape="Location", color="Reef")
```

A minor discontinuity was in north reefs samples were observed but the results were similar.


```{r}
q2_admixture_unlinked <- read_table("data/hpc/pca/atenius.ld_pruned_snps.pcangsd.admix.2.Q", col_names = c("X2","X1")) %>%
  add_column(sample_id = sample_ids) %>% 
  gather(cluster, proportion, -sample_id) %>% 
  left_join(sample_table) %>% 
  mutate(pop_order = site_order()[pop])

p1 <- ggplot(q2_admixture_unlinked ,aes(x=reorder(sample_id,pop_order),y=proportion)) + 
  geom_bar(aes(fill=cluster,color=cluster),stat="identity") + 
  facet_grid(reorder(location,desc(pop_order))~., scales = "free_y") + coord_flip() + theme_pubclean() + labs(x="",y="Proportion (K=2)") + 
  scale_fill_brewer(palette = "Set2") + 
  scale_color_brewer(palette = "Set2") + 
theme(axis.text.y = element_blank(),
        legend.position = "none", 
        axis.ticks = element_blank(),
        panel.border = element_blank(), 
        panel.background = element_blank(),
        panel.grid.minor = element_blank(),
        strip.text = element_text(size=12))
p1
```
