---
title: "Investigating potential batch effects"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.retina = 2)
library(tidyverse)
library(ggpubr)
source("functions.R")

```

All samples in this study were sequenced with the same sequencing technology, same read length and at approximately the same overall depth. In addition, as shown in analyses of [population structure](04.population_structure.md) no difference between inshore and offshore samples was observed in PCA or in a tree based on IBS distances.  Nevertheless we did observe some consistent differences between inshore and offshore samples that might indicate a subtle batch effect. 

The first of these differences was in individual heterozygosity estimates. To calculate individual heterozygosities we first ran `angsd` for each sample individually. Running `realSFS` on these individual allele frequency files provides an estimate of the number of heterozygous sites (AF=0.5) and we can obtain the heterozygosity by dividing this by the total number of sites. To a first approximation the total number of sites is simply equal to the number of sites with AF=0. 

```bash
for sample in samples;do
angsd -i $i -ref ${ref} -anc ${ref} -C 50 \
    -GL 2 -doSaf 1 \
    -sites ../reference_mappability_K50_E2.unique_mdust1M_depth.bed \
    -doCounts 1 \
    -setMinDepth 2 -setMaxDepth 12\
    -minQ 30 -minMapQ 30 -nThreads 15 -uniqueOnly 1 -doMajorMinor 1  -out $sample

    ../../angsd/misc/realSFS -fold 1 -P 15 ${sample}.saf.idx > ${sample}.ml 
done
```

```bash
for f in *.ml;do s=${f%.ml}; h=$(awk '{print $2/$1}' $f); printf "%s\t%s\n" $s $h;done > hets.tsv
```

We performed this individual heterozygosity calculation for our standard dataset as well as for data with coverage downsampled to 2x so that all samples have the same overall depth. 

```{r}
hets <- read_tsv("data/3.summary_ic/het_allsites/all_hets.tsv",col_names = c("sample","het"),show_col_types = F) %>% 
  sample2reef() %>% 
  reef2pop()

hets_ds <- read_tsv("data/3.summary_ic/het_downsample/all_hets.tsv",col_names = c("sample","het"),show_col_types = F) %>%   
  sample2reef() %>% 
  reef2pop() %>% 
  dplyr::rename(het_2x=het)

hets_gs <- read_tsv("data/3.summary_ic/het_goodwindows/goodwindows_hets.tsv",col_names = c("sample","het"),show_col_types = F) %>%   
  sample2reef() %>% 
  reef2pop() %>% 
  dplyr::rename(het_gs=het)

hetcov_data <- read_tsv("data/3.summary_ic/coverage.tsv",col_names = c("sample","cov"), show_col_types = F) %>% 
  left_join(hets) %>% 
  add_reef_order() %>% 
  left_join(hets_ds) %>% 
  left_join(hets_gs) %>% 
  pivot_longer(cols=c("het","het_2x","het_gs"),names_to = "dataset",values_to = "het")
```


```{r}
het.labs <- c("Full Data", "Downsampled to 2x")
names(het.labs) <- c("het", "het_2x")

hetcov_data %>% 
  filter(dataset!="het_gs") %>% 
  ggplot(aes(x=reorder(reef,reef_order),y=het)) + 
  geom_boxplot(aes(color=reorder(reef,reef_order)),outlier.shape = NA) + 
  geom_jitter(width = 0.3,size=0.5,aes(color=reorder(reef,reef_order))) + 
  scale_color_manual(values = site_colors(), labels=site_labels()) +
  facet_wrap(~dataset,ncol = 1,labeller = labeller(dataset = het.labs)) + 
  theme_pubr() + ylab("Heterozygosity %") + xlab("") + theme(legend.title = element_blank())
ggsave("figures/Figure_S7.png",width = 8,height = 6)
```


